---
title: "Recoding"
author: "Barry DeCicco"
date: "4/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load_libraries}
library(tidyverse)
library(magrittr)
```


## Recoding variables

In the tidyverse, the **dplyr** package is used for a lot of 
'data wrangling' tasks, such as creating new variables.


```{r create_data}
Mother <- c(0,0,0,1,1,NA)
Father <- c(0,1,1,0,0,1)
df <- data.frame(Mother, Father)
df
str(df) #both Mother and Father columns are numeric

```

## Mutate

With dply's **mutate** function, you can create new variables, or modify old ones:

```{r mutate1}

df <- df %>% 
  dplyr::mutate(combined = Mother + Father)

df
```
## Mutate - summarise

With dply's **mutate** function, and the summarise function,
you can create summary variables:

```{r mutate2}

df <- df %>% 
  dplyr::mutate(mean_combined = mean(combined))

df
```

#### Propagation of missing values:

NA's (missing values) **propagate**, meaning that functions will return NA if any of the values are NA.
```{r mutate3}

df <- df %>% 
  dplyr::mutate(mean_combined = mean(combined, na.rm=TRUE))

df
```


#### Recoding into New Variables:

Recoding variables is very common, and the **case_when** statement offers a simple method of doing this (in place of IF-ELSE statements inside of each other).  Below is an example of 

```{r case_when1}

df <- df %>% 
  mutate(new_variable = case_when(
    Mother == 0 & Father == 0 ~ 0, #condition 1
    Mother == 0 & Father == 1 ~ 1, #condition 2
    Mother == 1 & Father == 0 ~ 2, #condition 3
    Mother == 1 & Father == 1 ~ 3 #condition 4
    )
)#all other cases

df
```

#### Explicitly dealing with missing values:

In the case above, we did not account for any other combinations, such as either 'Mother' or 'Father' being missing.  We let the unhandled combinations default to missing.

Here, we explicitely declare that:



```{r case_when2}

df <- df %>% 
  mutate(new_variable = case_when(
    Mother == 0 & Father == 0 ~ 0, #condition 1
    Mother == 0 & Father == 1 ~ 1, #condition 2
    Mother == 1 & Father == 0 ~ 2, #condition 3
    Mother == 1 & Father == 1 ~ 3, #condition 4
    TRUE~ NA_real_                      #all other cases
    )
)

df

```


## References:
This code was borrowed from: https://stackoverflow.com/posts/60086150/revisions  <p>
Example of case_when:  https://www.rdocumentation.org/packages/dplyr/versions/0.7.8/topics/case_when  <p>
Documentation of case_when in tidyverse.org:
https://dplyr.tidyverse.org/reference/case_when.html  <p>
Documentation of dplyr  in tidyverse.org: 
https://dplyr.tidyverse.org/index.html <p>


