---
title: "File 1- EDA"
author: "Barry DeCicco"
date: "3/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What this is about

This file covers the initial 'poking and prodding' of a data set (in R, a 'data frame', or 'tibble').

The initial goals are to figure out the general nature of the data set.  The detailed statistical analyses come later.

The classic situations are:
1. You've imported a data set from somewhere else, and you want to see if the import process messed anything up, and 
2. Somebody handed you an R data frame (already imported), and you want to get an overall sense of what you are dealing with.

#### Library

This file will also serve as a collection of several different methods, which do overlapping tasks.  In some cases, the choice of method is based on what gives the nicest looking output.


## Libraries and data sets

The first step is to load the packages and data sets used in this file:

- **janitor**, 
- **kableExtra** 
- **inspectdf**
- **skimr**
- **magrittr**
- **gridExtra**   # for a grid layouts
- **DataExplorer**
- **htmltools**   # for inspectdf plots
- **tidyverse**


```{r libraries, echo=FALSE}
library(janitor)
library(kableExtra)
library(inspectdf)
library(skimr)
library(magrittr)
library(gridExtra)
library(DataExplorer)
# if you can not install 'DataExplorer' in the normal way, pull it from github:
# devtools::install_github("boxuancui/DataExplorer")
#  install.packages("htmltools") run the first time you run this script.
library(htmltools)

library(tidyverse)

```

#### Data sets:

- **mtcars**, 
- **mpg** 
- **iris**


```{r data_sets, echo=FALSE}

data(mtcars)
data(mpg)
data(iris)

```


## Why Do Exploratory Data Analysis?

- **Make Sure that the import went well**
- **Find out what data you have - by column**
- **Find out what data you have - by row**

**The first point is not emphasized, but should be - when importing data sets, it's really easy to have things not be understood correctly by R, resulting in numeric/date variables being read in as strings, or missing value codes not being understood.**


#### Make Sure that the import went well.

- **Is the created object a data frame?**
- **What are the row/column counts? (what did you expect?)**
- **What are the column types?  (Numeric/Character/Dates as expected)**
<br>
<br>
```{r Class_Dimension, echo=TRUE}

print(class(mpg))
print(dim(mpg))

```
Find out what the columns are like:

```{r Column_types, echo=TRUE}

names(mpg)
str(mpg)
```

Glimpse:  use the glimpse() command to get another look at the data set.  This is more compact than str(), but with less data.

```{r glimpse, echo=TRUE}

glimpse(mpg)


```


Get a list of the types (classes) of each column in a nicer arrangement:

```{r Column_Types_Better, echo=TRUE}

options(width = 80)

# From:  https://stackoverflow.com/questions/40306280/how-to-transpose-a-dataframe-in-tidyverse

df_column_types <- lapply(mpg, class) %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column() %>%  
  pivot_longer(-rowname) %>% 
  pivot_wider(names_from=rowname, values_from=value) %>%
  rename("Column_Name" = "name", "Column_Type" = "1")


grid.table(df_column_types)
```

#### Summaries within columns

When you want to see things by column:
1. Summary statistics for numeric data.
2. How many unique values are in each column?
3. How many missing values are in each column?

```{r Summary_1, echo=TRUE}

summary_df <- summary(mpg)  %>% as.data.frame()

summary_df

```
Unique values by column - the lapply function
will go over the columns of the first argument
(the data frame 'mpg') and apply the function given in the second argument ('unique')

```{r Summary_2, echo=TRUE}

 lapply(mpg, unique) 


```
Missing values - you want to know how many
missing values are in each column.  

R uses the 'NA' symbol to mean 'missing'.
It is not the actual letters 'NA', but rather
a symbol. This method uses the package 
'purr', from the tidyverse, to 'map' a function across the columns.

The function used is an improvised function, which applies the standard function 'is.na()'
to each element of the column, then sums them.
is.na() returns TRUE/FALSE for each elment;
when they are summed:
  TRUE is evaluated as 1;
  FALSE is evaluated as 0.

```{r Summary_3, echo=TRUE}

# from:  https://sebastiansauer.github.io/sum-isna/   'Way 2: using purrr:map'

missing_counts_by_column <- map(mpg,
                           ~sum(is.na(.))) %>% 
              as.data.frame()

missing_counts_by_column

```

Counting missing values by row.
Sometimes you want to track missing values 
by row of the data frame.

this uses the apply function, with a 'MARGIN'
argument to cause the application to be by rows.  'MARGIN=2' would apply the function by columns, producing a by-column count as in the previous example.


```{r Summary_4, echo=TRUE}

# from:  https://sebastiansauer.github.io/sum-isna/   'Way 4: Counting NAs rowwise using apply'

missing_counts_by_row <-apply(mtcars, 
              MARGIN = 1,
              function(x) sum(is.na(x)))   %>% 
          as.data.frame()


missing_counts_by_row 

```


## Using packages for EDA

### skimr  
(https://cran.r-project.org/web/packages/skimr/vignettes/skimr.html)

This is the first pacakage that we'll look at.
Think of skimr as a collection of useful functions.  One of those is to produce summaries.  Skimr produces different summaries for character and numeric columns, so its 
convenient to run it separately.  We will use
functions from the dplyr package to split
the mpg data sets into the subsets of character and numeric columns, and run skimr on each.

Skimr selects the appropriate summaries for character variables:

```{r skimr_character, echo=TRUE}

 mpg %>%  
  dplyr::select_if(is.character) %>% 
  skim()



```

Skimr selects the appropriate summaries for numeric variables:

```{r skimr_numeric, echo=TRUE}

mpg %>%  
  dplyr::select_if(is.numeric) %>% 
  skim()

```



```{r Summary_5, echo=TRUE}

 mpg %>%  
  dplyr::select_if(is.character) %>% 
  skim()


```

#### DataExplorer
(https://www.littlemissdata.com/blog/simple-eda)

This package runs a full report, in a new html document, covering a long list of items.

```{r create_report, echo=TRUE}
DataExplorer::create_report(mpg)

```

#### inspectdf
(https://www.rdocumentation.org/packages/inspectdf/versions/0.0.9.
https://www.littlemissdata.com/blog/inspectdf)
https://www.littlemissdata.com/blog/simple-eda



This package, like skimr, has a collection of functions which specialize in one task:

* inspect_na:  Summary and comparison of the rate of missingness across dataframe columns
* inspect_imb:  Summary and comparison of the most common levels in categorical columns
* inspect_cat:  Summary and comparison of the levels in categorical columns
* show_plot 	Simple graphical inspection of dataframe summaries
* inspect_types:  Summary and comparison of column types
* inspect_num:  Summary and comparison of numeric columns

Only two functions will be illustrated:
inspect_cat() and inspect_ibm(), both of which examine categorical variables.

```{r inspect_cat, echo=TRUE}

inspectdf::inspect_cat(mpg)

```

```{r inspect_imb, echo=TRUE}

inspectdf::inspect_imb(mpg)

```