---
title: "File 1- EDA"
author: "Barry DeCicco"
date: "3/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What this is about

This file covers the initial 'poking and prodding' of a data set (in R, a 'data frame', or 'tibble').

The initial goals are to figure out the general nature of the data set.  The detailed statistical analyses come later.

The classic situations are:
1. You've imported a data set from somewhere else, and you want to see if the import process messed anything up, and 
2. Somebody handed you an R data frame (already imported), and you want to get an overall sense of what you are dealing with.

#### Library

This file will also serve as a collection of several different methods, which do overlapping tasks.  In some cases, the choice of method is based on what gives the nicest looking output.


## Libraries and data sets

The first step is to load the packages and data sets used in this file:

- **janitor**, 
- **kableExtra** 
- **inspectdf**
- **skimr**
- **tidyverse**


```{r libraries, echo=FALSE}
library(janitor)
library(kableExtra)
library(inspectdf)
library(skimr)

library(tidyverse)

```

#### Data sets:

- **mtcars**, 
- **mpg** 
- **iris**


```{r data_sets, echo=FALSE}

data(mtcars)
data(mpg)
data(iris)

```



