---
title: "Tables in R"
author: "Barry DeCicco"
date: "4/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tables in R

This document shows the basics of making tables in R.
It starts with the basic methodds, which produce crude looking results, and then works upwards to better looking tables.


The first step is to load the packages and data sets used in this file:

- **janitor**, 
- **kableExtra** 
- **inspectdf**
- **skimr**
- **magrittr**
- **gridExtra**   # for a grid layouts
- **DataExplorer**
- **htmltools**   # for inspectdf plots
- **tidyverse**


```{r libraries, echo=FALSE}
library(janitor)
library(kableExtra)
library(inspectdf)
library(skimr)
library(magrittr)
library(gridExtra)

# if you can not install 'DataExplorer' in the normal way, pull it from github:
# devtools::install_github("boxuancui/DataExplorer")
#  install.packages("htmltools") run the first time you run this script.
library(htmltools)

library(tidyverse)

```


#### Data sets.  This uses some build-in data sets.
You can change to your own by substituting your data set name and the appropriate column names in the commands below.

- **mtcars**, 
- **mpg** 
- **iris**


```{r data_sets, echo=FALSE}

data(mtcars)
data(mpg)
data(iris)

```




## Tables

#### Basic table using the **table()** command.  

The default is not to show missing values, so add the option useNA = ifany".

Using two columns produces a two-way table;
Usinging three columns produces a three-way table.

The order is:  row variable, column variable, 
page variable.

```{r table1, echo=TRUE}

table(mtcars$cyl, useNA = "ifany")

table(mtcars$cyl, mtcars$gear, useNA = "ifany")
table(mtcars$cyl,mtcars$cyl,mtcars$carb, useNA = "ifany")
```

#### Proportions using the **prop.table()** command:

This command will take a table, and compute the proportions.  The ' %>%' is a pipe, which feeds the results of the line.  The round() command rounds it to the number of decimals which you specify.

(the '.' in the round command is just a placeholder, because you are using it in a pipe)


```{r table2, echo=TRUE}

table(mtcars$cyl, mtcars$gear, useNA = "ifany") %>% 
  prop.table() %>% 
  round(.,2)

```

Simplifying it a bit - you con't need to keep rewriting the data set name.  You can use another type of pipe ( %$%) or the 'with()' command.

```{r empty}


with(mtcars, table(cyl, gear, useNA = "ifany"))

mtcars %$% 
  table(cyl, gear, useNA = "ifany") %>% 
  prop.table() %>% 
  round(.,2)

```
#### Using the **tabyl()** function from the Janitor package  (and the %>% pipe).  The tabyl package has an **adorn** command to add features.  The adorn functions are almost self-explanatory.

```{r tabyl1}


mtcars %>% 
  janitor::tabyl(cyl, gear) 

mtcars %>% 
  janitor::tabyl(cyl, gear) %>% 
  adorn_totals("row") %>% 
  adorn_totals("col") %>% 
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns()

```

#### The kableExtra packages will  make tables look much better.  They also come with styling options.

kbl(caption = "Recreating booktabs style table") %>%
  kable_classic(full_width = F, html_font = "Cambria")



```{r KableExta1}


T1 <- mtcars %>% 
  janitor::tabyl(cyl, gear) 

  kableExtra::kbl(T1)

T2 <- mtcars %>% 
  janitor::tabyl(cyl, gear) %>% 
  adorn_totals("row") %>% 
  adorn_totals("col") %>% 
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns() 

  kbl(T2) %>% 
    kable_styling(bootstrap_options = c("striped", "hover"),
                  full_width = F)

  


```


## References:

https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

https://bookdown.org/kdonovan125/ibis_data_analysis_r4/working-with-tables-in-r.html

https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html

https://stackoverflow.com/questions/44528173/using-table-in-dplyr-chain