---
title: "Testing Data Set Opening"
author: "Barry DeCicco"
date: "4/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### This is the best description of what and why:  http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r load_libraries}
library(tidyverse)
library(magrittr)
library(here)
```
#### Create a data set and name it 'df':

```{r create_data}
Mother <- c(0,0,0,1,1,NA)
Father <- c(0,1,1,0,0,1)
df <- data.frame(Mother, Father)
df
str(df) #both Mother and Father columns are numeric

```

#### Demonstrate that we can print something:

The mean value of 'Father' is:  `r mean(df$Father)`

#### Save the data set to the hard drive as '...Data/df.Rdata':

```{r save_data}
saveRDS(df, file="Data/df.Rds", compress = F)

```

#### delete the data set from memory - it still exists on the hard drive, but is no longer loaded:

```{r delete_data}

rm(df)

```

#### Demonstrate that the data set 'df' is no longer available:

(the stuff being printed out is R telling us how to use 'str()'; it assumes that we are ignorant)

```{r test_delete_data}

str(df)

```




